Progress Report on NHANES Data

### 2023-12-19 Docs/NhanesApplExample.Rmd
slight modification
try with three PCs

# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
K <- 4
source(here("Code/OutsampBayes.R"))
```

@@

Because step 4 (debias) is very time-consuming, this example is fit on a subset of NHANES dataset, including 500 participants as the training set

@@

@@

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
# keep 4 PCs
t2 <- Sys.time()
# t2-t1 
```

```{r PC}
data.frame(t = mid, fpca_mod$efunctions[, 1:4]) %>%
  rename(PC1 = X1, PC2=X2, PC3=X3,PX4=X4) %>%
  pivot_longer(2:5, names_to = "PC") %>%
  ggplot()+
  geom_line(aes(x=t, y=value, col = PC))+
  labs(x="Time", y="", title = "Eigenfunctions from FPCA on the binned grid")
```

@@

```{r}
df_pc1 <- data.frame(t=1:J,  df_phi) %>%
  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>%
  pivot_longer(2:5, names_to = "PC")

df_pc2 <- data.frame(t = mid, fpca_mod$efunctions) %>%
  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>%
  pivot_longer(2:5, names_to = "PC")

@@

@@

t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(sind, bs="cc")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
# t2-t1
```

I ran into a few problems at this step. Not only does the mgcv::bam model take a really long time to fit, the re-evaluated eigenvalues are not proportional to the FPCA eigenvalues. They also lost the decreasing nature. Specifically:

- The smoothing parameters from the mgcv::bam model: `r debias_glmm$sp[2:5]`
- The re-evaluated eigenvalues, which are inverse of the smoothing parameters above: `r 1/debias_glmm$sp[2:5]`
- The **biased** eigenvalues from FPCA (step 3): `r fpca_mod$evalues[1:4]`

The FPCA eigenvalues, as expected, decreases from PC1 to PC4. However after re-evaluation, the eigenvalue corresponding to PC2 is less than PC3. Also, the proportion of these two set of estimates is not constant, as below:

```{r, class.source = 'fold-show'}
# proportion of biased eigenvalues to re-evaluated eigenvalues
fpca_mod$evalues/(1/debias_glmm$sp[2:5])
```

>>>>

# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
K <- 3
source(here("Code/OutsampBayes.R"))
```

@@

Because step 4 (debias) is very time-consuming, this example is fit on a subset of NHANES dataset, including 1000 participants as the training set

@@

Follow up on the email yesterday, we would like to see what would happen if we only take the first **three** PC functions.

@@

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
# keep 3 PCs
t2 <- Sys.time()
# t2-t1

# fpca_mod$evalues
```

```{r PC}
data.frame(t = mid, fpca_mod$efunctions) %>%
  rename(PC1 = X1, PC2=X2, PC3=X3) %>%
  pivot_longer(2:4, names_to = "PC") %>%
  ggplot()+
  geom_line(aes(x=t, y=value, col = PC))+
  labs(x="Time", y="", title = "Eigenfunctions from FPCA on the binned grid")
```

@@

```{r}
df_pc1 <- data.frame(t=1:J,  df_phi) %>%
  rename(PC1=X1, PC2=X2, PC3=X3) %>%
  pivot_longer(2:4, names_to = "PC")

df_pc2 <- data.frame(t = mid, fpca_mod$efunctions) %>%
  rename(PC1=X1, PC2=X2, PC3=X3) %>%
  pivot_longer(2:4, names_to = "PC")

@@

Again, here we use the three PC functions in the random effect part.

@@

t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(sind, bs="cc")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re"),
                     # s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
# t2-t1 # 3.5 minutes
```

It seems that the eigenvalues are still losing their decreasing nature. The eigenvalue corresponding to PC2 is still smaller than PC3. Specifically

- The smoothing parameters from the mgcv::bam model: `r debias_glmm$sp[2:4]`
- The re-evaluated eigenvalues, which are inverse of the smoothing parameters above: `r 1/debias_glmm$sp[2:4]`
- The **biased** eigenvalues from FPCA (step 3): `r fpca_mod$evalues[1:3]`

Also, the proportion of these two set of estimates is not constant, as below:

```{r, class.source = 'fold-show'}
# proportion of biased eigenvalues to re-evaluated eigenvalues
fpca_mod$evalues/(1/debias_glmm$sp[2:4])
```

### 2023-12-19 Docs/ProgressReportAppl.Rmd
try with three PCs

- Original sample size took too much time in the debias step of fGFPCA method. 
- In fact, the debais step does not work very well in this case
- 500 subjects for training, 500 for out-of-sample prediction

@@

Ntr <- Nte <- 500

train_id <- unique(df$id)[1:500]
test_id <- unique(df$id)[501:1000]

train_df <- df %>% filter(id %in% train_id)
test_df <- df %>% filter(id %in% test_id)
# 1440*500
```

@@

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
t2 <- Sys.time()
t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model

# dim(fpca_mod$efunctions) # 27 eigenfunctions
K <- 4
```

@@

t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(sind, bs="cc")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
t2-t1
```

>>>>

- 60% subjects for training, 40% for out-of-sample prediction
- Original sample size took too much time in the debias step of fGFPCA method. 
- In fact, the debais step does not work very well in this case

@@

Ntr <- round(N*0.6)
Nte <- N - Ntr

train_id <- sample(unique(df$id), Ntr)
train_df <- df %>% filter(id %in% train_id)

test_df <- df %>% filter(!id %in% train_id)
test_id <- unique(test_df$id)

# 1440*500
```

@@

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
t2 <- Sys.time()
t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model
```

@@

t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(sind, bs="cc")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
t2-t1 # 20 hours

save(debias_glmm, train_df, test_df,
      file = here("Data/Appl_debias_model.RData"))
